---
title: "name of assignment"
author: "student name"
date: "date"
output:
  pdf_document
header-includes:
- \renewcommand*\familydefault{\sfdefault} %% this picks a sans serif font
- \usepackage[T1]{fontenc}
---
```{r setup, echo=F}
knitr::opts_chunk$set(cache = F)
```

**DELETE ANYTHING FROM THIS TEMPLATE BELOW THAT IS NOT PART OF YOUR SOLUTION.**

(0) Instructions for installing tinytex for PDF rendering: [**https://yihui.org/tinytex/**](https://yihui.org/tinytex/)
 
```{r, eval = F}
install.packages('tinytex')
tinytex::install_tinytex()
```

### I. First section

(1) Answer to first question...using `mono-spaced font` to refer to a `function()` or other `code`...

```{r}
## code for first question...
a <- 1:4
a
```

(2) Here's a way to make a plot and include it

```{r, fig.width=5, fig.height=4}
plot(1:10)
```

### II. Second section

(3) Make sure you read about `cache`-ing: https://bookdown.org/yihui/rmarkdown-cookbook/cache.html

(4) If you can't get your R Markdown document to render as a PDF, you can try rendering it to HTML, and then "printing" the .html file to a PDF using any internet browser. 

